A Horrocks-type theorem for even orthogonal $K_2$

### 2019-09-28 dl-paper.tex
Fix 5.19

\subsection{The subgroups $P_\alpha(0)$, $P_\alpha(*)$ and their properties.} \label{sec:Pa0-basic}
For a root system $\Phi$ consider the following subsets of $\Phi$:
\begin{align} Z_+(\alpha) & = \{ \beta \in \Phi \mid \langle \alpha, \beta \rangle > 0 \}, \\
   Z_0(\alpha) & = \{ \beta \in \Phi \mid \alpha + \beta \not\in \Phi,\ \langle \alpha, \beta \rangle = 0 \}, \\
   Z(\alpha)   & = Z_0(\alpha) \sqcup Z_+(\alpha). \end{align}
Clearly, $Z_0(\alpha)$ (resp. $Z_+(\alpha)$) is a reductive (resp. special) subset of $\Phi$.

@@

\subsection{The subgroups $K(\alpha, \beta)$ and their properties} \label{sec:Pa0-Steinberg}
Throughout this subsection $\alpha, \beta$ denote a fixed pair of roots of $\Phi$ forming a sharp angle (i.\,e. $\langle \alpha, \beta \rangle = 1$). We denote by $\Psi$ the subsystem of type $\rA_2$ generated by $\alpha$ and $\beta$.
\begin{df}\label{def:Kab} Denote by $\widetilde{K}(\alpha, \beta)$ the subgroup of $\St(\Phi, A[X], M[X])$ generated by the following five families of elements (below $\xi \in A[X]$, $f\in M[X]$):
 \begin{enumerate}
  \item $z_\gamma(Xf, \xi)$, for all $\gamma \in \Phi \setminus \Psi$;
  \item $x_{-\alpha}(X^2f)$, $x_{-\beta}(X^2f)$;
  \item $x_{\alpha}(Xf)$, $x_\beta(Xf)$;
  \item $x_{\alpha-\beta}(Xf)$, $x_{\beta-\alpha}(Xf)$;
  \item $c_{\delta}(f, X\xi)$ for some fixed root $\delta \in \Phi \setminus \Psi$. \end{enumerate}
Denote by $K(\alpha, \beta)$ the image of $\widetilde{K}(\alpha, \beta)$ in $\St(\Phi, A[X, X\inv])$. \end{df}

\begin{prop} \label{K-a-b} Every element $g \in K(A[X], M[X])$ can be written as a product $g_0 \cdot x_{-\alpha}(mX) \cdot x_{-\beta}(m'X)$ for some $m, m' \in M$ and $g_0 \in \widetilde{K}(\alpha, \beta)$. 
Consequently, $K(\alpha, \beta)$ coincides with the intersection $\mathrm{Im}(K(A[X], M[X]) \to \St(\Phi, A[X, X\inv])) \cap P_\alpha(0) \cap P_\beta(0)$. In particular, $K(\alpha, \beta)$ does not depend on the choice of the root $\delta$. \end{prop}
\begin{proof} 
For $i=1..5$ denote by $\mathcal{G}_i$ the set of all generators of type $i$ from~\cref{def:Kab}.
By~\cref{thm:Stepanov} every element of $\St(\Phi, A[X], XM[X])$ can be presented as a product of generators $\mathcal{G}_1$ and generators $\mathcal{G}' = \{ x_{\gamma}(Xf) \mid \gamma \in \Psi\}$. In turn, by~\cref{Kgen-strong} one can express $g$ as a product of these generators and generators $\mathcal{G}_5$. %We are going to rewrite this product so that the generators of $\widetilde{K}(\alpha, \beta)$ accumulate on the left-hand side of this expression, while the elements of the root subgroups $X_{-\alpha}(MX)$ and $X_{-\beta}(MX)$ accumulate on the right-hand side.

By~\cref{thm:Stepanov} $\widetilde{K}(\alpha, \beta)$ contains the normal subgroup $G_0 = \St(\Phi, A[X], X^2M[X])$, so it suffices to obtain the required presentation of $g$ modulo $G_0$. By~\cref{Zrels} and~\cref{Crels} the elements of $\mathcal{G}'$ commute with the elements of $\mathcal{G}_1$ and $\mathcal{G}_5$ modulo $G_0$, therefore we can rewrite $g$ as $g_1 \cdot g_2$, where $g_1 \in \langle \mathcal{G}_1, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_2 \in \langle \mathcal{G}' \rangle$.

Now let us study the structure of the subgroup $\langle \mathcal{G}' \rangle$. It is clear that  $x_{\gamma_1}(mX)$ and $x_{\gamma_2}(m'X)$ commute modulo $G_0$ whenever $\gamma_1 \neq - \gamma_2$, $\gamma_1, \gamma_2 \in \Psi$. On the other hand, for $\gamma \in \Psi$ any commutator $c_{\gamma}(mX, m'X)$ is congruent to some generator of $\mathcal{G}_5$  modulo $G_0$ (cf. the proof of~\cref{Kgen-strong}) and, therefore, commutes with the elements of $\mathcal{G}'$ modulo $G_0$.

Now pick each factor $x_{\gamma}(Xf)$, $\gamma \in \{ -\alpha, -\beta \}$ appearing in the presentation for $g_2$, decompose it as $x_{\gamma}(X^2f') \cdot x_{\gamma}(m'X)$ for some $f', m'$ and then move the second factor to the rightmost position within $g_2$ conjugating all factors along the way. By the previous paragraph only elements of  $\mathcal{G}_5$ may appear modulo $G_0$ after the simplification of these conjugates.
Thus, we can rewrite $g_2$ as $g_{21} \cdot g_{22}$, where $g_{21} \in \langle \mathcal{G}_2, \mathcal{G}_3, \mathcal{G}_4, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_{22} \in \langle X_{-\alpha}(MX), X_{-\beta}(MX) \rangle = X_{-\alpha}(MX) \cdot X_{-\beta}(MX)$.
Thus, we have obtained the desired decomposition for $g$.

It is clear from the definition that $K(\alpha, \beta) \subseteq P_\alpha(0) \cap P_\beta(0) \cap K(A[X], M[X])$. The reverse inclusion follows from the just proved decomposition and the fact that for $g\in K(A[X], M[X])$ one has $p_{\alpha}(g) = m$ and $p_\beta(g) = m'$ (cf.~\cref{rem:palpha}). \end{proof}

@@

\begin{lemma} \label{image-K-a-b} The image of $K(\alpha, \beta)$ under $S_{\alpha}(a, -)$ is contained in the subgroup of $\St(\Phi, A[X, X^{-1}])$  generated by $K(\alpha, \beta)$ and $Z_{\alpha, \beta}$. \end{lemma}
\begin{proof} To simplify notation we call the generators of the subgroup $Z_{\alpha,\beta}$ ``generators of type Z``. By~\eqref{eq:Zconj1}, \eqref{eq:Zconj2} the conjugate to each generator $z_\gamma(Xf, \xi)$ of type 1 by $x_\alpha(aX^{-1})$ is a product of generators of type Z, Z, 1.
By~\eqref{eq:Zconj3} the conjugate to the generator $x_{-\alpha}(X^2f)$ is a product of generators of type 1, Z, 1, 2, 1, 1, Z, Z\@.
The conjugation by $x_\alpha(aX^{-1})$ fixes both generators of type 3 and transforms $x_{-\beta}(X^2f)$ into a product of generators of type 2, 1. As for the remaining generators of type 5, \cref{Crels} shows that the corresponding conjugate can be written as a product of generators of type Z, Z, 5. \end{proof}

>>>>

\subsection{The subgroups $P_\alpha(0)$, $P_\alpha(*)$ and their properties.} \label{sec:Pa0-basic}
Let $\Phi$ be a root system and $\alpha \in \Phi$ be its element.
Consider the following subsets of $\Phi$:
\begin{align} Z_+(\alpha) & = \{ \beta \in \Phi \mid \langle \alpha, \beta \rangle > 0 \}, \\
   Z_0(\alpha) & = \{ \beta \in \Phi \mid \alpha + \beta \not\in \Phi,\ \langle \alpha, \beta \rangle = 0 \}, \\
   Z(\alpha)   & = Z_0(\alpha) \sqcup Z_+(\alpha). \end{align}
Clearly, $Z_0(\alpha)$ (resp. $Z_+(\alpha)$) is a reductive (resp. special) subset of $\Phi$.

@@

\subsection{The subgroups $K(\alpha, \beta)$ and their properties} \label{sec:Pa0-Steinberg}
Throughout this subsection $\alpha, \beta$ denote a fixed pair of roots of $\Phi$ forming a sharp angle (i.\,e. $\langle \alpha, \beta \rangle = 1$). We denote by $\Psi$ the subsystem of type $\rA_2$ generated by $\alpha$ and $\beta$.
\begin{df}\label{def:Kab} Denote by $\widetilde{K}(\alpha, \beta)$ the subgroup of $\overline{\St}(\Phi, A[X], M[X])$ generated by the following five families of elements parametrized by $\xi \in A[X]$, $f\in M[X]$:
 \begin{enumerate}
  \item $z_\gamma(Xf, \xi)$, for all $\gamma \in \Phi \setminus \Psi$;
  \item $x_{-\alpha}(X^2f)$, $x_{-\beta}(X^2f)$;
  \item $x_{\alpha}(Xf)$, $x_\beta(Xf)$;
  \item $x_{\alpha-\beta}(Xf)$, $x_{\beta-\alpha}(Xf)$;
  \item $c_{\delta}(f, X\xi)$ for some fixed root $\delta \in \Phi \setminus \Psi$. \end{enumerate}
Denote by $K(\alpha, \beta)$ the image of $\widetilde{K}(\alpha, \beta)$ under $j_+ \colon \St(\Phi, A[X]) \to \St(\Phi, A[X, X\inv])$. \end{df}

\begin{prop} \label{K-a-b} Every element $g \in K(A[X], M[X])$ can be written as a product $g_0 \cdot x_{-\alpha}(mX) \cdot x_{-\beta}(m'X)$ for some $m, m' \in M$ and $g_0 \in \widetilde{K}(\alpha, \beta)$. 
Consequently, $K(\alpha, \beta)$ coincides with the intersection $j_+(K(A[X], M[X])) \cap P_\alpha(0) \cap P_\beta(0)$. In particular, $K(\alpha, \beta)$ does not depend on the choice of the root $\delta$. \end{prop}
\begin{proof} 
For $i=1,\ldots,5$ denote by $\mathcal{G}_i$ the set of all generators of type $i$ from~\cref{def:Kab}.
By~\cref{thm:Stepanov} every element of $\St(\Phi, A[X], XM[X])$ can be presented as a product of generators $\mathcal{G}_1$ and generators $\mathcal{G}' = \{ x_{\gamma}(Xf) \mid \gamma \in \Psi\}$. In turn, by~\cref{Kgen-strong} one can express $g$ as a product of these generators and generators $\mathcal{G}_5$.

By~\cref{thm:Stepanov} $\widetilde{K}(\alpha, \beta)$ contains normal subgroup $G_0 = \overline{\St}(\Phi, A[X], X^2M[X])$, so it suffices to obtain the required presentation of $g$ modulo $G_0$. By~\cref{Zrels} and~\cref{Crels} the elements of $\mathcal{G}'$ commute with the elements of $\mathcal{G}_1$ and $\mathcal{G}_5$ modulo $G_0$, therefore we can rewrite $g$ as $g_1 \cdot g_2$, for some $g_1 \in \langle \mathcal{G}_1, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_2 \in \langle \mathcal{G}' \rangle$.

It is clear that  $x_{\gamma_1}(mX)$ and $x_{\gamma_2}(m'X)$ commute modulo $G_0$ whenever $\gamma_1 \neq - \gamma_2$, $\gamma_1, \gamma_2 \in \Psi$. On the other hand, for $\gamma \in \Psi$ any commutator $c_{\gamma}(mX, m'X)$ is congruent to some generator of $\mathcal{G}_5$  modulo $G_0$ (cf. the proof of~\cref{Kgen-strong}) and hence commutes with the elements of $\mathcal{G}'$ modulo $G_0$.

Now pick each factor $x_{\gamma}(Xf)$, $\gamma \in \{ -\alpha, -\beta \}$ appearing in the presentation for $g_2$, decompose it as $x_{\gamma}(X^2f') \cdot x_{\gamma}(m'X)$ for some $f', m'$ and then move the second factor to the rightmost position within $g_2$ conjugating all factors along the way. By the previous paragraph only elements of  $\mathcal{G}_5$ may appear modulo $G_0$ after simplification of these conjugates.
Thus, we can rewrite $g_2$ as $g_{21} \cdot g_{22}$, where $g_{21} \in \langle \mathcal{G}_2, \mathcal{G}_3, \mathcal{G}_4, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_{22} \in \langle X_{-\alpha}(MX), X_{-\beta}(MX) \rangle = X_{-\alpha}(MX) \cdot X_{-\beta}(MX)$.
Thus, we have obtained the desired decomposition for $g$.

It is clear from the definition that $K(\alpha, \beta) \subseteq P_\alpha(0) \cap P_\beta(0) \cap j_+(K(A[X], M[X]))$. The reverse inclusion follows from the just proved decomposition and the fact that for $g\in j_+(K(A[X], M[X]))$ one has $p_{\alpha}(g) = m$ and $p_\beta(g) = m'$ (cf.~\cref{rem:palpha}). \end{proof}

@@

\begin{lemma} \label{image-K-a-b} The image of $K(\alpha, \beta)$ under conjugation by $x_\alpha(aX^{-1})$ is contained in the subgroup of $\St(\Phi, A[X, X^{-1}])$  generated by $K(\alpha, \beta)$ and $Z_{\alpha, \beta}$. \end{lemma}
\begin{proof} To simplify notation we call the generators of the subgroup $Z_{\alpha,\beta}$ ``generators of type Z``. By~\eqref{eq:Zconj1}, \eqref{eq:Zconj2} the conjugate to each generator $z_\gamma(Xf, \xi)$ of type 1 by $x_\alpha(aX^{-1})$ is either the generator itself or a product of generators of type Z, Z, 1. \cref{Crels} also shows that the conjugate to a generator of type 5 is either the generator itself or a product of generators of type Z, Z, 5.

The conjugation by $x_\alpha(aX^{-1})$ fixes both types of generators of type 3 and transforms $x_{-\beta}(X^2f)$ into a product of generators of type 2, 4. The conjugation also fixes $x_{-\alpha-\beta}(Xf)$ and transforms $x_{\beta-\alpha}(Xf)$ into a product of generators of type 4, Z.

Now pick a root $\gamma \in \Phi \setminus \Psi$ such that $\alpha + \gamma \in \Phi$.
From~\eqref{eq:Zconj3} in which $\gamma$ is substituted for $\beta$ it follows that  $z_{-\alpha}(X^2f, aX^{-1})$ is a product of generators of type 1, Z, 1, 2, 1, 1, Z, Z\@.
\end{proof}
